quollr: An R Package for Visualizing $2-D$ Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-05-07 paper-quollr.Rmd
Updated sections

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. We also want to load the built-in data sets `s_curve_noise_training` and `s_curve_noise_umap`.

`s_curve_noise_training` is a $3\text{-}D$ S-curve data set with additional four noise dimensions which is used to train the model. `s_curve_noise_umap` is the UMAP $2\text{-}D$ embedding for `s_curve_noise_training` data set. Each data set contains a unique ID column.

### Scaling the data

The algorithm begins by scaling NLDR data to a standard scale using the `gen_scaled_data()` function. This function standardizes the data with first embedding scale to 0 and 1, The second embedding scales to 0 and the maximum of the second embedding. The resulting output provides the scaled NLDR data and the original limits of embeddings.

@@

We can compute the `r2` which is the ratio of the ranges of the original embedding components.

@@

## Constructing the $2\text{-}D$ model

Constructing the $2\text{-}D$ model mainly contains (i) binning data, (ii) obtaining bin centroids, and (iii) indicating neighbors by line segments connecting centroids.

>>>>

## Data sets

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. We also want to load the built-in data sets `scurve` and `scurve_umap`.

`scurve` is a $3\text{-}D$ S-curve data set with additional four noise dimensions which is used to build the model. `scurve_umap` is the UMAP $2\text{-}D$ embedding for `scurve` data set with `n_neighbors` is $15$ and `min_dist` is $0.1$. Each data set contains a unique ID column that mapping `scurve` and `scurve_umap`.

### Scaling the Data

The algorithm starts by scaling the NLDR data to a standard range using the `gen_scaled_data()` function. This function standardizes the data so that the first embedding ranges from $0$ to $1$, while the second embedding scales from $0$ to the maximum value of the second embedding. The output includes the scaled NLDR data along with the original limits of the embeddings.

@@

We can calculate the `r2`, which represents the ratio of the ranges of the original embedding components.

@@

## Constructing the $2\text{-}D$ Model

Constructing the $2\text{-}D$ model primarily involves (i) binning data, (ii) obtaining bin centroids, and (iii) connecting centroids with line segments to indicate neighbors.
